### 2023-04-20 paper.tex
typos etc

This procedure is commonly adapted when working with single-cell DNA methylation data, because once on gets to the PCA step, one can then continue with the established methods just mentioned. The task is therefore to construct from the methylation data a matrix suitable as input for PCA.
A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g. 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes the as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
This matrix is now subjected to PCA.
After PCA, one can proceed with dimensionality reduction and clustering approaches known from scRNA-seq.

@@

%\phantomsection
\subsection{Read-position aware quantitation} \label{residuals}

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval. Typically, read coverage per cell is sparse in scBS data. In the example of Figure \ref{fig:smoothres}A, teh depicted interval is covered by a single read for two of the three cells shown and no read in the third. The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 1 to have higher methylation in the interval than cell 2.
However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

@@

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{figures/Fig_correlation.png}
	\end{center}
	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promotors, correlation is weak. Promoters are defined as intervals \textpm2~kb around the TSS.
		\textbf{(B)} Mean methylation near the gene Htra1.
		Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
		Data from \citet{kremer_scnmt}.}
	\label{fig:correlation}
\end{figure*}

@@

We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promotors (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation (Fig.~\ref{fig:correlation}B).
One such example gene, Htra1, is depicted in Fig.~\ref{fig:correlation}C.
While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high Htra1 expression.

@@

We next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
To this end, we obtained cell type/state labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
To do this, we subjected the methylomes to four variants of the workflow described above: we defining genomic intervals in two ways and quantified methylation in these intervals in two ways, then always subjected the resulting matrix to PCA.
To define the intervals, we either simply tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages (``classic'' pre-processing), or we calculated the shrunken mean of residuals.

Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
Furthermore, even cellular sub-types of cells in the continuous neural stem cell lineage were partially separated.

@@

Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our ``scbs'' software, default 2000 )and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
This parameter sweep showed that our workflow gives good results over a wide range of parameter values. For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B). In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times. Overall, however, our default parameter combination provided good results and fast compute times in both data sets.

\begin{figure*}
	\begin{center}
		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
	\end{center}
	\caption{\small \textbf{Identification of differentially methylated regions.}\\
		The chromosome is divided into windows overlapping by a small and fixed size.
		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
		Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
	\label{fig:dmrscan}
\end{figure*}

@@

Similar to the previously described approach for VMR identification (Fig.~\ref{fig:vmr}), we divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
Next, instead of the variance, we obtain the t-statistic % Welch t-test
as a measure of differential methylation between the two cell groups.
We identify the windows with the most extreme t-statistics, e.g. windows in the 2\% upper and lower tails. We the merge any overlapping windows in the upper tail into larger DMRs, and do the same for windows in the lower tail, and then we re-calculate the t-statistic for each larger DMR.

To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels. The DMR t-statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.

@@

Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
Our results suggest that oligodendrocyte-specific gene expression of Mbp is supported by low methylation at the detected DMR.

@@

The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering. For this, the methylation matrix should in lieu of the log-normalized expression matrix usually expected by these tools.

@@

where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I. \todo{No Bessel correction?}

@@

We also calculate a variance as in Eq.(\ref{var}):
$$ v^g_I = \frac{1}{|C_\text{g}|+1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
\qquad g=\text{A},\text{B}$$
\todo{Is it really not pooled? Is there a Bessel correction?}

From this, we calculate Welch's t statistic as usual:
$$ t_I = \frac{\mu^\text{B}_I - \mu^\text{A}_I}{\sqrt{v^\text{B}_I - v^\text{A}_I}}.$$

In order to find candidate DMRs, we again define overlapping and stepwise shifted intervals $I_1, I_2, \dots$ as for the VMRs and calulate a t statistics $t_1, t_2, \dots$ for these. As before, we take the top 2-percentile of these values, fuse intervals that overlap and call the regions thus obtained candidate DMRs. We repeat the procedure for the bottom 2-percentile to get the candidate DMRs for the other sign.

Next, we need to check these candidate DMRs for statistical significance. We first remind the readers here that, as this is a within-sample analysis, and cells, not samples, are the statistical unit. Therefore, a call as significant implies that the same DMR is likely to be called again if we repated the analysis with another set of cells taken from the same biological sample, not that it would generalize to further samples. This fact, although often overlooked, is common to all within-sample analyses in the single-cell fields, e.g. also to the differential expression tests performed in scRNASeq analyses to find marker genes that differentiate clusters.

It may seem that we could use the standard procedure for the Welch t-test here, i.e. use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution. However, this is unlikely to hold for two reasons: First, the Welch-Sattersthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.  Second, the fusing of the DMRs obtained in the scanning step to obtain fused candidate DMRs would invalidate subsequent p-value based adjustment for multiple testing.

Therefore, we have instead implemented a permutation procedure, which works as follows: We randomly shuffle the assignment of the cells in $C_\text{A}\cup C_\text{A}$ to either of the two groups and then rerun the whole procedure, i.e., the scanning step, the DMR fusing and the calculation of t values from the (potentially fused) candidate DMRs. This needs to be done for a sufficiently large number of permutation. Running through the whole genome for each permutation would be too computationally expensive. Instead, we go through the genome only once \todo{correct?}, but reshuffle the group labels every \todo{???} kb.

All the t values obtained from this permutation procedure are taken together to get an empirical null distibution. Then, we can use this null distribution to control false discovery rate (FDR) by applying the Benjamini-Hochberg procedure in its p-value-free form: Let us write $T$ for the set of all t values obtained from the ``real'' assignment of cells to group labels and $T_0$ from the set of all t values obtained from the shuffled assignments, i.e., the empirical null distribution. To adjust a specific t value $t_i\in T$, we calculate
$$ p^\text{adj}_i = \frac{ \left|\left\{t_j\in T_0\mid|t_j| > |t_i| \right\}\right| \big/ |T_0|}
{ \left|\left\{t_j\in T\mid|t_j| > |t_i| \right\}\right| \big/ |T|}.$$
In words: we calculate which fraction of the null t values is greater than t by absolute value, and which fraction of the real t values is. The ratio gives us the false discovery rate we should expect if we used the t value as threshold. \todo{Check if this is what we do.}

@@

\setcounter{figure}{0}
%\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\thefigure}{S\arabic{figure}}

>>>>

This procedure is commonly adapted when working with single-cell DNA methylation data, because once on gets to the PCA step, one can then continue with the established methods just mentioned. The task is therefore to construct from the methylation data a matrix suitable as input for PCA.
A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of e.g. 100~kb size, and calculate for each cell the average methylation of the DNA within each tile.
To this end, one identifies in the tile all CpG sites that are covered by at least one read and averages their methylation state, i.e., one denotes as average DNA methylation of the tile in a given cell the proportion of the observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A).
This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (``methylation fractions'') between 0 and 1.
This matrix is now subjected to PCA.
After PCA, one can proceed with dimensionality reduction and clustering approaches known from scRNA-seq.

@@

\subsection{Read-position aware quantitation} \label{residuals}

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval. Typically, read coverage per cell is sparse in scBS data. In the example of Figure \ref{fig:smoothres}A, the depicted interval is covered by a single read for two of the three cells shown and no read in the third. The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider cell 1 to have higher methylation in the interval than cell 2.
However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval.
Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

@@

@@

We first note that gene expression is more strongly correlated with the methylation of nearby VMRs than with the methylation of their promotors (Fig.~\ref{fig:correlation}A), indicating that VMR methylation is often a better predictor of gene expression than promoter methylation.
Indeed, a gene-wise comparison revealed many genes whose expression is correlated with the nearest VMR, but not with promoter methylation.
One such example gene, \textit{Htra1}, is depicted in Fig.~\ref{fig:correlation}C.
While the promoter of this gene is lowly methylated regardless of gene expression, a VMR located downstream of the promoter is lowly methylated in cells with high \textit{Htra1} expression.

@@

We next tested whether our methods improve the ability to distinguish cell types and cell states (Fig.~\ref{fig:score}).
To this end, we obtained cell type/state labels based on the single-cell transcriptomes from \citet{kremer_scnmt} (Fig.~\ref{fig:score}A).
We consider these cell labels as ground truth and tested whether we are able to distinguish the same groups of cells based on their methylomes.
To do this, we subjected the methylomes to four variants of the workflow described above: we defined genomic intervals in two ways and quantified methylation in these intervals in two ways, then always subjected the resulting matrix to PCA.
To define the intervals, we either simply tiled the genome into 100~kb bins as is commonly done, or we used our VMR detection approach.
To quantify methylation in these intervals, we either averaged in these intervals, obtaining methylation percentages (``classic'' pre-processing), or we calculated the shrunken mean of residuals.

Visual inspection of the resulting UMAPs revealed that our proposed combination of methods results in more clearly separated cell types, compared to a UMAP obtained with default analysis methods (Fig.~\ref{fig:score}B,C).
While all cells form a continuous point cloud when using default methods, our improvements led to a clear separation of oligodendrocytes, OPCs and endothelial cells.
Furthermore, even cellular sub-states of cells in the continuous neural stem cell lineage were partially separated.

@@

Lastly, we assessed whether our proposed workflow requires fine-tuning of parameters (Fig.~\ref{fig:sweep}).
To this end, we re-analyzed both data sets, as well as sub-samples of the data with different VMR detection parameters, namely the width of the sliding window in bp, (set with option \texttt{--bandwidth} in our ``scbs'' software, default 2000) and the variance threshold above which windows are merged to VMRs (\texttt{--var-threshold}, default 0.02).
This parameter sweep showed that our workflow gives good results over a wide range of parameter values. For the data of \citet{luo2017single}, results are nearly independent of the parameters (Fig.~\ref{fig:sweep}B). In the more challenging data set of \citet{kremer_scnmt} (Fig.~\ref{fig:sweep}A), cell types were less cleanly separated when very large bandwidths, or very strict variance thresholds were selected.
However, very small bandwidths or very lenient thresholds resulted in a much higher number of VMRs and thus long computing times. Overall, however, our default parameter combination provided good results and fast compute times in both data sets.

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{figures/Fig_correlation.png}
	\end{center}
	\caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
		\textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red): for most promotors, correlation is weak. Promoters are defined as intervals \textpm2~kb around the TSS.
		\textbf{(B)} Mean methylation near the gene \textit{Htra1}.
		Cells are assigned to 4 groups based on \textit{Htra1} expression (group 0: cells with no \textit{Htra1} expression, groups 1-3: cells that express \textit{Htra1}, divided into three equally large groups with group 3 having the highest expression).
		Data from \citet{kremer_scnmt}.}
	\label{fig:correlation}
\end{figure*}

@@

\begin{figure*}
	\begin{center}
		\includegraphics[width=.75\textwidth]{figures/Fig_DMRdetection.png}
	\end{center}
	\caption{\small \textbf{Identification of differentially methylated regions.}\\
		The chromosome is divided into windows overlapping by a small and fixed size.
		For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
		Then the t statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
		Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t statistic above or below the threshold.
		Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
	\label{fig:dmrscan}
\end{figure*}

Similar to the previously described approach for VMR identification (Fig.~\ref{fig:vmr}), we divide each chromosome into overlapping windows shifted by a small and fixed size (step size) and quantify the methylation of each cell in each window.
Next, instead of the variance, we obtain the t statistic % Welch t-test
as a measure of differential methylation between the two cell groups.
We identify the windows with the most extreme t statistics, e.g. windows in the 2\% upper and lower tails. We then merge any overlapping windows in the upper tail into larger DMRs, and do the same for windows in the lower tail, and then we re-calculate the t statistic for each larger DMR.

To assess statistical significance of DMRs, we repeat the same procedure on permutations of the scBS data, i.e. the same data set with randomly shuffled cell labels. The DMR t statistics obtained from permuted data are then used to estimate the false-discovery rate (FDR), yielding an adjusted p-value for each DMR.

@@

Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (\textit{Mbp}), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
Our results suggest that oligodendrocyte-specific gene expression of \textit{Mbp} is supported by low methylation at the detected DMR.

@@

The methylation matrix produces by the \texttt{matrix} function can then serve as input for standard single-cell analysis tools like Seurat \citep{seurat} or scanpy \citep{Wolf_2018}, in order to perform, e.g., dimensionality reduction and cell clustering. For this, the methylation matrix should be provided in lieu of the log-normalized expression matrix usually expected by these tools.

@@

where the average runs only over the set $C_I=\bigcup_{i\in I}C_i$ of those cells $j$ which have reads covering interval I.

@@

We also calculate a variance as in Eq.(\ref{var}):
$$ v^g_I = \frac{1}{|C_\text{g}|-1}  \sum_{j\in g} \left( r_{Ij} - \mu_I^g \right)^2, 
\qquad g=\text{A},\text{B}$$

From this, we calculate Welch's t statistic as usual:
$$ t_I = \frac{\mu^\text{A}_I - \mu^\text{B}_I}{\sqrt{v^\text{A}_I/|C_A| + v^\text{B}_I/|C_B|}}.$$

In order to find candidate DMRs, we again define overlapping and stepwise shifted intervals $I_1, I_2, \dots$ as for the VMRs and calulate t statistics $t_1, t_2, \dots$ for these. As before, we take the top 2-percentile of these values, fuse intervals that overlap and call the regions thus obtained candidate DMRs. We repeat the procedure for the bottom 2-percentile to get the candidate DMRs for the other sign.

Next, we need to check these candidate DMRs for statistical significance. We first remind the readers here that, as this is a within-sample analysis, and cells, not samples, are the statistical unit. Therefore, a call as significant implies that the same DMR is likely to be called again if we repeated the analysis with another set of cells taken from the same biological sample, not that it would generalize to further samples. This fact, although often overlooked, is common to all within-sample analyses in the single-cell fields, e.g. also to the differential expression tests performed in scRNA-seq analyses to find marker genes that differentiate clusters.

It may seem that we could use the standard procedure for the Welch t-test here, i.e. use the Welch-Satterthwaite formula to get an approximate degree of freedom and then calculate the tail probability of the corresponding t distribution. However, this is unlikely to hold for two reasons: First, the Welch-Satterthwaite degrees of freedom are only based on the number of cells per group and do not account for the fact that the read coverage might vary from cell to cell.  Second, the fusing of the DMRs obtained in the scanning step to obtain fused candidate DMRs would invalidate subsequent p-value based adjustment for multiple testing.

Therefore, we have instead implemented a permutation procedure, which works as follows: We randomly shuffle the assignment of the cells in $C_\text{A}\cup C_\text{A}$ to either of the two groups and then rerun the whole procedure, i.e., the scanning step, the DMR fusing and the calculation of t values from the (potentially fused) candidate DMRs. This needs to be done for a sufficiently large number of permutations. Running through the whole genome for each permutation would be too computationally expensive. Instead, we go through the genome only once, but reshuffle the group labels every 2 megabasepairs.  % didn't abbreviate cause one of our example genes is "Mbp"...

All the t values obtained from this permutation procedure are taken together to get an empirical null distibution. Then, we can use this null distribution to control false discovery rate (FDR) by applying the Benjamini-Hochberg procedure in its p-value-free form: Let us write $T$ for the set of all t values obtained from the ``real'' assignment of cells to group labels and $T_0$ from the set of all t values obtained from the shuffled assignments, i.e., the empirical null distribution. To adjust a specific t value $t_i\in T$, we calculate
$$ p^\text{adj}_i = \frac{ \left|\left\{t_j\in T_0\mid|t_j| > |t_i| \right\}\right| \big/ |T_0|}
{ \left|\left\{t_j\in T\mid|t_j| > |t_i| \right\}\right| \big/ |T|}.$$

\todo{In the code we do, for a given t-stat threshold: adjPval = fdisc / (fdisc + tdisc). This seems slightly different than the formula here.}

In words: we calculate which fraction of the null t values is greater than t by absolute value, and which fraction of the real t values is. The ratio gives us the false discovery rate we should expect if we used the t value as threshold.

@@

\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}
